---
title: "SWong Data 605 Homework 3"
author: "Susanna Wong"
date: "2023-09-12"
output: 
  html_document:
    theme: cerulean
    toc: true
    toc_float: true
editor_options: 
  markdown: 
    wrap: sentence
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Assignment Prompt

Problem set 1

(1) What is the rank of the matrix A?

A = $\left[\begin{array}{ccc} 1 & 2 & 3 & 4\\ -1 & 0 & 1 & 3\\ 0 & 1 & -2 & 1\\ 5 & 4 & -2 & -3\\ \end{array}\right]$

(2) Given an mxn matrix where m \> n, what can be the maximum rank?
    The minimum rank, assuming that the matrix is non-zero?

(3) What is the rank of matrix B?

B = $\left[\begin{array}{ccc} 1 & 2 & 1\\ 3 & 6 & 3\\ 2 & 4 & 2\\ \end{array}\right]$

Problem set 2

Compute the eigenvalues and eigenvectors of the matrix A.
You'll need to show your work.
You'll need to write out the characteristic polynomial and show your solution.

A = $\left[\begin{array}{ccc} 1 & 2 & 3\\ 0 & 4 & 5\\ 0 & 0 & 6\\ \end{array}\right]$

Please show your work using an R-markdown document.

# Problem Set 1 {.tabset}

## Question 1 {.tabset}

What is the rank of the matrix A?

A = $\left[\begin{array}{ccc} 1 & 2 & 3 & 4\\ -1 & 0 & 1 & 3\\ 0 & 1 & -2 & 1\\ 5 & 4 & -2 & -3\\ \end{array}\right]$

### Show work

First, we find the row echelon form of matrix A.

**Step 1**

$\left[\begin{array}{cccc} 1 & 2 & 3 & 4\\ -1 & 0 & 1 & 3\\ 5 & 4 & -2 & -3 \end{array}\right]\underrightarrow{R_{2}=R_{1}+R_{2}}\left[\begin{array}{cccc} 1 & 2 & 3 & 4\\ 0 & 2 & 4 & 7\\ 5 & 4 & -2 & -3 \end{array}\right]$

**Step 2**

$-5R_{1}=\left[\begin{array}{cccc} -5 & -10 & -15 & -20\end{array}\right]$

$\left[\begin{array}{cccc} 1 & 2 & 3 & 4\\ 0 & 2 & 4 & 7\\ 5 & 4 & -2 & -3 \end{array}\right]\underrightarrow{R_{3}=-5R_{1}+R_{3}}\left[\begin{array}{cccc} 1 & 2 & 3 & 4\\ 0 & 2 & 4 & 7\\ 0 & -6 & -17 & -23 \end{array}\right]$

**Step 3**

$3R_{2}=\left[\begin{array}{cccc} 0 & 6 & 12 & 21\end{array}\right]$

$\left[\begin{array}{cccc} 1 & 2 & 3 & 4\\ 0 & 2 & 4 & 7\\ 0 & -6 & -17 & -23 \end{array}\right]\underrightarrow{R_{3}=3R_{2}+R_{3}}\left[\begin{array}{cccc} 1 & 2 & 3 & 4\\ 0 & 2 & 4 & 7\\ 0 & 0 & -5 & -2 \end{array}\right]$

Since there are three nonzero rows in the reduced the matrix A, **the rank of Matrix A is 3.**

### R Code

```{r}
A <- matrix(c(1, -1, 5, 2, 0, 4, 3, 1, -2, 4, 3, -3),nrow = 3)
A
```

```{r}
library(Matrix)
rankMatrix(A)
```

## Question 2

Given an mxn matrix where m \> n, what can be the maximum rank?
The minimum rank, assuming that the matrix is non-zero?

According to Stat Trek, "the **rank** of a matrix is defined as (a) the maximum number of [linearly independent](https://stattrek.com/statistics/dictionary?definition=Linear_dependence_of_vectors) *column* vectors in the matrix" If m \> n, the maximum rank of the matrix is n.
If the matrix does not contain any elements, the rank of the matrix is 0.
If the matrix contains at least 1 element, the minimum rank of the matrix is 0.

[Source: Stat Trek](https://stattrek.com/matrix-algebra/matrix-rank#:~:text=Since%20the%20matrix%20has%20more,number%20of%20linearly%20independent%20rows.)

## Question 3 {.tabset}

What is the rank of matrix B?

B = $\left[\begin{array}{ccc} 1 & 2 & 1\\ 3 & 6 & 3\\ 2 & 4 & 2\\ \end{array}\right]$

### Without Calculation

We have a 3x3 matrix.
The maximum feasible rank of the matrix is the minimum of the dimension, which is 3.
Since column 1 and column 3 are the same, one of them has to go.
Column 2 is a multiple of column 1.
Row 2 and 3 are each multiples of row 1.
The rank of matrix B is 1.

### Show work

Step 1

$-3R_{1}=\left[\begin{array}{ccc} -3 & -6 & -3\end{array}\right]$

$\left[\begin{array}{ccc} 1 & 2 & 1\\ 3 & 6 & 3\\ 2 & 4 & 2 \end{array}\right]\underrightarrow{R_{2}=-3R_{1}+R_{2}}\left[\begin{array}{ccc} 1 & 2 & 1\\ 0 & 0 & 0\\ 2 & 4 & 2 \end{array}\right]$

**Step 2**

$-2R_{1}=\left[\begin{array}{ccc} -2 & -4 & -2\end{array}\right]$

$\left[\begin{array}{ccc} 1 & 2 & 1\\ 0 & 0 & 0\\ 2 & 4 & 2 \end{array}\right]\underrightarrow{R_{3}=-2R_{1}+R_{3}}\left[\begin{array}{ccc} 1 & 2 & 1\\ 0 & 0 & 0\\ 0 & 0 & 0 \end{array}\right]$

Since there are one nonzero rows in the reduced the matrix B, **the rank of Matrix B is 1.**

### R Code

```{r}
B <- matrix(c(1, 3, 2, 2, 6, 4, 1, 3, 2),nrow = 3)
B
```

The rank of matrix B is 1.

```{r}
rankMatrix(B)
```

# Problem Set 2 {.tabset}

Compute the eigenvalues and eigenvectors of the matrix A.
You'll need to show your work.
You'll need to write out the characteristic polynomial and show your solution.

A = $\left[\begin{array}{ccc} 1 & 2 & 3\\ 0 & 4 & 5\\ 0 & 0 & 6\\ \end{array}\right]$

Please show your work using an R-markdown document.

## Eigenvalues {.tabset}

### Show work

$det(\lambda I-A)=0$ $\lambda I=\lambda\left[\begin{array}{ccc} 1 & 0 & 0\\ 0 & 1 & 0\\ 0 & 0 & 1 \end{array}\right]=\left[\begin{array}{ccc} \lambda & 0 & 0\\ 0 & \lambda & 0\\ 0 & 0 & \lambda \end{array}\right]$

$det\left(\left[\begin{array}{ccc} \lambda & 0 & 0\\ 0 & \lambda & 0\\ 0 & 0 & \lambda \end{array}\right]-\left[\begin{array}{ccc} 1 & 2 & 3\\ 0 & 4 & 5\\ 0 & 0 & 6 \end{array}\right]\right)=0$

$det\left(\left[\begin{array}{ccc} \lambda-1 & -2 & -3\\ 0 & \lambda-4 & -5\\ 0 & 0 & \lambda-6 \end{array}\right]\right)=0$

Row 3 has the most zeros.
Expand on row 3.

$0\left|\begin{array}{cc} -2 & -3\\ \lambda-4 & -5 \end{array}\right|-0\left|\begin{array}{cc} \lambda-1 & -3\\ 0 & -5 \end{array}\right|+(\lambda-6)\left|\begin{array}{cc} \lambda-1 & -2\\ 0 & \lambda-4 \end{array}\right|=0$

$[0]-[0]+(\lambda-6)[(\lambda-1)(\lambda-4)-(0)(-2)]=0$

$(\lambda-6)[(\lambda-1)(\lambda-4)-0]=0$

$(\lambda-6)(\lambda-1)(\lambda-4)=0$

$\lambda=$ 1, 4, or 6

### R Code

```{r}
A <- matrix(c(1, 0, 0, 2, 4, 0, 3, 5, 6), nrow = 3 )
A
```

```{r}
eigen_A <- eigen(A) 
eigen_A$values
```

```{r}
```

```{r}
```

## Eigenvectors {.tabset}

### lambda = 1

$\lambda=1$

$\left[\begin{array}{ccc} \lambda-1 & -2 & -3\\ 0 & \lambda-4 & -5\\ 0 & 0 & \lambda-6 \end{array}\right]=\left[\begin{array}{ccc} 1-1 & -2 & -3\\ 0 & 1-4 & -5\\ 0 & 0 & 1-6 \end{array}\right]=\left[\begin{array}{ccc} 0 & -2 & -3\\ 0 & -3 & -5\\ 0 & 0 & -5 \end{array}\right]$

We find the reduced echelon form of the matrix above.

$\left[\begin{array}{ccc} 0 & -2 & -3\\ 0 & -3 & -5\\ 0 & 0 & -5 \end{array}\right]\underrightarrow{R_{1}=-\frac{R_{1}}{2}}\left[\begin{array}{ccc} 0 & 1 & \frac{3}{2}\\ 0 & -3 & -5\\ 0 & 0 & -5 \end{array}\right]$

$\left[\begin{array}{ccc} 0 & 1 & \frac{3}{2}\\ 0 & -3 & -5\\ 0 & 0 & -5 \end{array}\right]\underrightarrow{R_{2}=3R_{1}+R_{2}}\left[\begin{array}{ccc} 0 & 1 & \frac{3}{2}\\ 0 & 0 & -\frac{1}{2}\\ 0 & 0 & -5 \end{array}\right]$

$\left[\begin{array}{ccc} 0 & 1 & \frac{3}{2}\\ 0 & 0 & -\frac{1}{2}\\ 0 & 0 & -5 \end{array}\right]\underrightarrow{R_{2}=-2R_{2}}\left[\begin{array}{ccc} 0 & 1 & \frac{3}{2}\\ 0 & 0 & 1\\ 0 & 0 & -5 \end{array}\right]$

$\left[\begin{array}{ccc} 0 & 1 & \frac{3}{2}\\ 0 & 0 & 1\\ 0 & 0 & -5 \end{array}\right]\underrightarrow{R_{3}=5R_{2}+R_{3}}\left[\begin{array}{ccc} 0 & 1 & \frac{3}{2}\\ 0 & 0 & 1\\ 0 & 0 & 0 \end{array}\right]$

$\left[\begin{array}{ccc} 0 & 1 & \frac{3}{2}\\ 0 & 0 & 1\\ 0 & 0 & 0 \end{array}\right]\underrightarrow{R_{3}=\frac{3}{2}R_{2}+R_{1}}\left[\begin{array}{ccc} 0 & 1 & 0\\ 0 & 0 & 1\\ 0 & 0 & 0 \end{array}\right]$

To find the null space, we solve $\left[\begin{array}{ccc} 0 & 1 & 0\\ 0 & 0 & 1\\ 0 & 0 & 0 \end{array}\right]\left[\begin{array}{c} v_{1}\\ v_{2}\\ v_{3} \end{array}\right]=\left[\begin{array}{c} 0\\ 0\\ 0 \end{array}\right]$

We have $v_{2}=0;v_{3}=0$

$v_{\lambda=1}=\left[\begin{array}{c} 1\\ 0\\ 0 \end{array}\right]$

### lambda = 4

$\lambda=4$

$\left[\begin{array}{ccc} \lambda-1 & -2 & -3\\ 0 & \lambda-4 & -5\\ 0 & 0 & \lambda-6 \end{array}\right]=\left[\begin{array}{ccc} 4-1 & -2 & -3\\ 0 & 4-4 & -5\\ 0 & 0 & 4-6 \end{array}\right]=\left[\begin{array}{ccc} 3 & -2 & -3\\ 0 & 0 & -5\\ 0 & 0 & -2 \end{array}\right]$

We find the reduced echelon form of the matrix above.

$\left[\begin{array}{ccc} 3 & -2 & -3\\ 0 & 0 & -5\\ 0 & 0 & -2 \end{array}\right]\underrightarrow{R_{1}=\frac{R_{1}}{3}}\left[\begin{array}{ccc} 1 & -\frac{2}{3} & -1\\ 0 & 0 & -5\\ 0 & 0 & -2 \end{array}\right]$

$\left[\begin{array}{ccc} 1 & -\frac{2}{3} & -1\\ 0 & 0 & -5\\ 0 & 0 & -2 \end{array}\right]\underrightarrow{R_{2}=-\frac{R_{2}}{5}}\left[\begin{array}{ccc} 1 & -\frac{2}{3} & -1\\ 0 & 0 & 1\\ 0 & 0 & -2 \end{array}\right]$

$\left[\begin{array}{ccc} 1 & -\frac{2}{3} & -1\\ 0 & 0 & 1\\ 0 & 0 & -2 \end{array}\right]\underrightarrow{R_{1}=R_{2}+R_{1}}\left[\begin{array}{ccc} 1 & -\frac{2}{3} & 0\\ 0 & 0 & 1\\ 0 & 0 & -2 \end{array}\right]$

$\left[\begin{array}{ccc} 1 & -\frac{2}{3} & 0\\ 0 & 0 & 1\\ 0 & 0 & -2 \end{array}\right]\underrightarrow{R_{3}=2R_{2}+R_{3}}\left[\begin{array}{ccc} 1 & -\frac{2}{3} & 0\\ 0 & 0 & 1\\ 0 & 0 & 0 \end{array}\right]$

$\left[\begin{array}{ccc} 1 & -\frac{2}{3} & 0\\ 0 & 0 & 1\\ 0 & 0 & 0 \end{array}\right]\left[\begin{array}{c} v_{1}\\ v_{2}\\ v_{3} \end{array}\right]=\left[\begin{array}{c} 0\\ 0\\ 0 \end{array}\right]$

$v_{1}-\frac{2}{3}v_{2}=0\Longrightarrow v_{1}=\frac{2}{3}v_{2}$

$v_{3}=0$

$v_{\lambda=4}=\left[\begin{array}{c} \frac{2}{3}\\ 1\\ 0 \end{array}\right]$

### lambda = 6

$\lambda=6$

$\left[\begin{array}{ccc} \lambda-1 & -2 & -3\\ 0 & \lambda-4 & -5\\ 0 & 0 & \lambda-6 \end{array}\right]=\left[\begin{array}{ccc} 6-1 & -2 & -3\\ 0 & 6-4 & -5\\ 0 & 0 & 6-6 \end{array}\right]=\left[\begin{array}{ccc} 5 & -2 & -3\\ 0 & 2 & -5\\ 0 & 0 & 0 \end{array}\right]$

We find the reduced echelon form of the matrix above.

$\left[\begin{array}{ccc} 5 & -2 & -3\\ 0 & 2 & -5\\ 0 & 0 & 0 \end{array}\right]\underrightarrow{R_{1}=\frac{R_{1}}{5}}\left[\begin{array}{ccc} 1 & -\frac{2}{5} & -\frac{3}{5}\\ 0 & 2 & -5\\ 0 & 0 & 0 \end{array}\right]$

$\left[\begin{array}{ccc} 1 & -\frac{2}{5} & -\frac{3}{5}\\ 0 & 2 & -5\\ 0 & 0 & 0 \end{array}\right]\underrightarrow{R_{2}=\frac{1}{2}R_{2}}\left[\begin{array}{ccc} 1 & -\frac{2}{5} & -\frac{3}{5}\\ 0 & 1 & -\frac{5}{2}\\ 0 & 0 & 0 \end{array}\right]$

$\left[\begin{array}{ccc} 1 & -\frac{2}{5} & -\frac{3}{5}\\ 0 & 1 & -\frac{5}{2}\\ 0 & 0 & 0 \end{array}\right]\underrightarrow{R_{1}=-\frac{2}{5}R_{2}+R_{1}}\left[\begin{array}{ccc} 1 & 0 & -\frac{8}{5}\\ 0 & 1 & -\frac{5}{2}\\ 0 & 0 & 0 \end{array}\right]$

$\left[\begin{array}{ccc} 1 & 0 & -\frac{8}{5}\\ 0 & 1 & -\frac{5}{2}\\ 0 & 0 & 0 \end{array}\right]\left[\begin{array}{c} v_{1}\\ v_{2}\\ v_{3} \end{array}\right]=\left[\begin{array}{c} 0\\ 0\\ 0 \end{array}\right]$

$v_{1}-\frac{8}{5}v_{3}=0\Longrightarrow v_{1}=\frac{8}{5}v_{3}$

$v_{2}-\frac{5}{2}v_{3}=0\Longrightarrow v_{2}=\frac{5}{2}v_{3}$

$v_{\lambda=6}=\left[\begin{array}{c} \frac{8}{5}\\ \frac{5}{2}\\ 1 \end{array}\right]$

### R Code

The eigenvectors does not match my calculations.
Perhaps it is normalized.

```{r}
eigen_A$vectors
```

Normalized eigenvectors = $\frac{\left[\begin{array}{c} v_{1}\\ v_{2}\\ \vdots\\ v_{n} \end{array}\right]}{\sqrt{v_{1}^{2}+v_{2}^{2}+\cdots+v_{n}^{2}}}$

```{r}
eigen_A$vectors
```

```{r}
norm_eigen_lambda1 <- matrix(c(1,0,0),nrow=3)/sqrt(1^2+0^2+0^2)
norm_eigen_lambda1
```

```{r}
norm_eigen_lambda4 <- matrix(c(2/3,1,0),nrow=3)/sqrt((2/3)^2+1^2+0^2)
norm_eigen_lambda4
```

```{r}
norm_eigen_lambda6 <- matrix(c(8/5,5/2,1),nrow=3)/sqrt((8/5)^2+(5/2)^2+1^2)
norm_eigen_lambda6
```

```{r}
norm_eigenvector <-cbind(norm_eigen_lambda6, norm_eigen_lambda4, norm_eigen_lambda1)
norm_eigenvector
```

Now we check if the our calculated normalized eigenvector matches with the output of the R code.
The first element in column one and column to do not match.
However, when displaying and comparing the eigenvectors, it appears to match.

```{r}
norm_eigenvector == eigen_A$vectors
```

```{r}
norm_eigenvector
```

```{r}
eigen_A$vectors
```

```{r}
norm_eigen_lambda0 <- matrix(c(-1,0,1),nrow=3)/sqrt(1^2+(-1)^2+0^2)
norm_eigen_lambda0
```

Upon [researching](https://zief0002.github.io/matrix-algebra/eigenvalues-and-eigenvectors.html), R does return the normalized eigenvector.

# Source

**Matrix Rank**

[Stat trek](https://stattrek.com/matrix-algebra/matrix-rank#){.uri}

**Normalized Eigenvectors**

[Matrix Algebra for Educational Scientist](https://zief0002.github.io/matrix-algebra/eigenvalues-and-eigenvectors.html){.uri}

[Eigenvector](https://byjus.com/jee/normalized-and-decomposition-of-eigenvectors/#){.uri}
